---
title: "Covid91 vaccine study Final2023"
author: "Dana Schroer"
output: 
  html_document:
    number_sections: true
    toc: true
    toc_depth: 5
date: "`r Sys.Date()`" 
---

```{r include=FALSE,echo=FALSE}
require(tidyverse)
require(tigerstats)
require(forcats)
FinalDat <-read.csv(file="final.csv",header=TRUE)
#recode to remove the annoying "X" from the dataframe
FinalData <- FinalDat  %>% select(-one_of("X"))
FinalData <- FinalData %>%
mutate(across(where(is.character), as.factor))
male = filter(FinalData, sex =="M")
female = filter(FinalData, sex =="F")
lgbtq = filter(FinalData, LGBTQ =="gay")
drugUser = filter(FinalData, DrugUser =="yes")
```

# Introduction

A new  preventative vaccine (covid921) is under test in clinical trials. Given the nature of this vaccine it is suspected that the drug may exhibit different effectiveness across the variety of subpopulations tracked. Consider population subgroups by gender( guys and gals), drug use, and sexual orientation. Write a complete report that separately considers this drug for each subgroup and your recommendations as to whether people in the subgroup should take the vaccine. Structure your report in four sections each devoted to one of the four subgroups, use null and alternative hypothesis statements and correct statistical verbage when referencing p-values in conclusions.

## Males

### Graphical Descriptive Results

```{r}
barchartGC(~infected+treatment,data=male)
barchartGC(~infected+treatment,data=male,type="percent")
```
  
  The bar chart above shows the graphical outcomes for males, based on the collected data. It compares the rates of contracting Covid-19 as compared to vaccination status. The left shows people who contracted Covid-19 and the right shows men who have not. This clearly shows that more than twice the percentage of people who are in the placebo group contracted Covid where people who did not contract Covid are very closely related.
  
### Numerical Descriptive Results

```{r}
male <- xtabs(~infected + treatment, data = male)
rowPerc(male)
colPerc(male)
```
 
  The set of data illustrates a twofold difference in individuals who received the vaccinations and encountered Covid-19. The placebo group contracted Covid approximately 70% of the time where the vaccinated group contracted Covid 30% of the time. However, there was no significant contrast among those who received vaccinations and were deemed "OK." We can see that the vaccine is generally effective in this testing scenario.
  
### Inferential Results

```{r}
chisq.test(male)
chisqtestGC(male)
maleFisher <- fisher.test(male)
maleFisher
```

  The Chi Squared test shows that the p-value for males is 0.0000000000001098. This is substantially smaller than the general use point of 0.05. Consequently, the very small number indicats that there is a difference in the data. The Fisher's Exact test yielded a p-value of 0.00000000000004631, signifying a significant difference in the occurrence of Covid-19 in the data. This all shows that the odds a unvaxinated man getting Covid are 2.2 times higher than a man with the vaccine contracting Covid.
  
## Females



## LGBTQ



## Druggies


# Overall Results and Conclusions