---
title: "Covid91 vaccine study Final2023"
author: "Dana Schroer"
output: 
  html_document:
    number_sections: true
    toc: true
    toc_depth: 5
date: "`r Sys.Date()`" 
---

```{r include=FALSE,echo=FALSE}
require(tidyverse)
require(tigerstats)
require(forcats)
FinalDat <-read.csv(file="final.csv",header=TRUE)
#recode to remove the annoying "X" from the dataframe
FinalData <- FinalDat  %>% select(-one_of("X"))
FinalData <- FinalData %>%
mutate(across(where(is.character), as.factor))
male = filter(FinalData, sex =="M")
female = filter(FinalData, sex =="F")
lgbtq = filter(FinalData, LGBTQ =="gay")
drugUser = filter(FinalData, DrugUser =="yes")
```

# Introduction

A new  preventative vaccine (covid921) is under test in clinical trials. Given the nature of this vaccine it is suspected that the drug may exhibit different effectiveness across the variety of subpopulations tracked. Consider population subgroups by gender( guys and gals), drug use, and sexual orientation. Write a complete report that separately considers this drug for each subgroup and your recommendations as to whether people in the subgroup should take the vaccine. Structure your report in four sections each devoted to one of the four subgroups, use null and alternative hypothesis statements and correct statistical verbage when referencing p-values in conclusions.

$H_A$ : There are statistically significant difference in the data, meaning that the vaccine and placebo are NOT equally as effective.

$H_0$ : There are no statistically significant difference in the data, meaning that the vaccine and placebo are equally as effective.

# Methods

Looking at the data collected, this will be categorical explained by categorical statistical problem, we will create separate data variables for each group to separate and test each of them. We will use the candP helper file to use R studios built in bar charts to display data graphically. W will also use xtabs to find the percentages relevant to the data. In this case we are only comparing two types of data and therefore we do not need to use any loops to measure multiple variables in the Fisher test or run a FisherXY.

## Males

### Graphical Descriptive Results

```{r}
barchartGC(~infected+treatment,data=male)
barchartGC(~infected+treatment,data=male,type="percent")
```
  
  The bar chart above shows the graphical outcomes for males, based on the collected data. It compares the rates of contracting Covid-19 as compared to vaccination status. The left shows people who contracted Covid-19 and the right shows men who have not. This clearly shows that more than twice the percentage of people who are in the placebo group contracted Covid where people who did not contract Covid are very closely related.
  
### Numerical Descriptive Results

```{r}
tableMale <- xtabs(~infected + treatment, data=male)
rowPerc(tableMale)
colPerc(tableMale)
```
 
  The set of data illustrates a twofold difference in individuals who received the vaccinations and encountered Covid-19. The placebo group contracted Covid approximately 70% of the time where the vaccinated group contracted Covid 30% of the time. However, there was no significant contrast among those who received vaccinations and were deemed "OK." We can see that the vaccine is generally effective in this testing scenario.
  
### Inferential Results

```{r}
chisq.test(male)
chisqtestGC(male)
maleFisher <- fisher.test(male)
maleFisher
```

  The Chi Squared test shows that the p-value for males is 0.0000000000001098. This is substantially smaller than the general use point of 0.05. Consequently, the very small number indicats that there is a difference in the data. The Fisher's Exact test yielded a p-value of 0.00000000000004631, signifying a significant difference in the occurrence of Covid-19 in the data. This all shows that the odds a unvaxinated man getting Covid are 2.2 times higher than a man with the vaccine contracting Covid.
  
### Confidence Interval

```{r}
oddsMale = maleFisher$estimate/(maleFisher$estimate+1)*100
print(oddsMale)
maleLow=(maleFisher$conf.int[1]/(maleFisher$conf.int[1]+1))*100
maleHigh=(maleFisher$conf.int[2]/(maleFisher$conf.int[2]+1))*100
cat("\n","Condifence interval for Males: ",maleLow," to ",maleHigh)
```

The confidence interval is >50% with a normal vaccine threshold of >60%. This confidence interval is more than adequate to show that this is a good vaccine for this group.

## Females

### Graphical Descriptive Results

```{r}
barchartGC(~infected + treatment,data=female)
barchartGC(~infected + treatment,data=female, type="percent")
```

Presented above is a bar chart that illustrates the graphical outcomes exclusively for females, based on the collected data. The comparison focuses on the rates of contracting Covid-19 relative to vaccination status, with the left side representing women who contracted the virus and the right side depicting women who did not. It is evident that approximately one-third more of the individuals in the placebo group contracted Covid-19, while the percentage of those who did not contract the virus remains closely aligned.

### Numerical Descriptive Results

```{r}
female <- xtabs(~infected + treatment, data=female)
rowPerc(female)
colPerc(female)
```

The numerical findings reinforce the insights gleaned from the graphical results. In the placebo group, Covid was contracted approximately 60% of the time, while the vaccinated group experienced a 40% contraction rate. It is apparent that the vaccine demonstrates general effectiveness in this testing scenario, albeit with a lower efficacy compared to male participants.

### Inferential Results
```{r}
chisq.test(female)
chisqtestGC(female)
femaleFisher <- fisher.test(female)
femaleFisher
```

  The results of the Chi-Squared test reveal a P-value for the tested data that is below the commonly accepted threshold of 0.05; in fact, it is 0.0002732, a small value signifying a significant difference in the data. The Chi-Squared statistic is 13, surpassing our degrees of freedom of one in this dataset. Additionally, the odds of an unvaccinated woman contracting Covid are 1.5 times higher than those of a woman with the vaccine.

### Confidence Interval

```{r}
oddsFemale = femaleFisher$estimate/(femaleFisher$estimate +1)*100
print(oddsFemale)
femaleLow=(femaleFisher$conf.int[1]/(femaleFisher$conf.int[1]+1))*100
femaleHigh=(femaleFisher$conf.int[2]/(femaleFisher$conf.int[2]+1))*100
cat("\n","Condifence interval for Females: ",femaleLow," to ",femaleHigh)
```

We see that the confidence interval is >50% and includes the 60% threshold, it is less effective here than in the male group but still a good vaccine.

## LGBTQ

### Graphical Descriptive Results

```{r}
barchartGC(~infected + treatment,data=lgbtq)
barchartGC(~infected + treatment,data=lgbtq, type="percent")
```

Displayed above is a bar chart illustrating the graphical outcomes exclusively for drug users derived from the dataset. The comparison involves the incidence of contracting Covid-19 based on vaccination status, with the left side depicting individuals who contracted Covid-19 and the right side representing those who did not. Once more, it is evident that a significantly higher number of individuals in the vaccinated group contracted Covid-19, while the non-contracted Covid-19 group shows a close correlation. Considering the sample size, the likelihood of this occurrence being solely due to chance is improbable.

### Numerical Results

```{r}
tableLGBTQ <- xtabs(~infected + treatment, data=lgbtq)
rowPerc(tableLGBTQ)
colPerc(tableLGBTQ)
```

The numerical results build on what the graphical results were showing. The placebo group contracted Covid approximately 60% of the time where the vaccinated group contracted Covid 40% of the time. We can see that the vaccine is generally effective in this testing scenario, although it is less effective then the male participants.

### Inferential Results

```{r}
chisq.test(tableLGBTQ)
chisqtestGC(tableLGBTQ)
lgbtqFisher <- fisher.test(tableLGBTQ)
lgbtqFisher
```

The Chi Squared test shows that the p-value for the tested data is much smaller than the general use point of 0.05. In fact, it is 0.00000006668, which indicates that there is a difference in the data. The Chi-Squared statistic is also 29, being larger than our degrees of freedom of one from this data. The odds of an unvaccinated person contracting Covid in this group are about 1:5.

### Confidence Interval

```{r}
oddsLGBTQ = lgbtqFisher$estimate/(lgbtqFisher$estimate +1)*100
print(oddsLGBTQ)
lgbtqLow=(lgbtqFisher$conf.int[1]/(lgbtqFisher$conf.int[1]+1))*100
lgbtqHigh=(lgbtqFisher$conf.int[2]/(lgbtqFisher$conf.int[2]+1))*100
cat("\n","Condifence interval for LGBTQ: ",lgbtqLow," to ",lgbtqHigh)
```

The confidence interval is 7-26% far lower than 50%. We can say that this is not a very good vaccine.

## Druggies

```{r}
barchartGC(~infected + treatment,data=drugUser)
barchartGC(~infected + treatment,data=drugUser, type="percent")
```

### Graphical Descriptive Results

```{r}
barchartGC(~infected + treatment,data=drugUser)
barchartGC(~infected + treatment,data=drugUser, type="percent")
```

The presented bar chart depicts the graphical outcomes for individuals identified as drug users and their likelihood of contracting Covid-19 based on their vaccination status. The left side indicates positive cases, while the right side shows negative cases. Notably, in this scenario, individuals who had Covid-19 were five to six times more likely to have received the vaccine rather than the placebo. This observation raises the possibility that drug users may have a heightened predisposition to contracting Covid-19, potentially due to weakened immune systems or a reverse effect of the vaccination when combined with drugs in a person's system. However, drawing definitive conclusions on this matter would necessitate a separate and focused study.

### Numerical Results

```{r}
tableDrug <- xtabs(~infected + treatment, data=drugUser)
rowPerc(tableDrug)
colPerc(tableDrug)
```

The numerical findings complement the insights derived from the graphical representation. In the placebo group, COVID contraction occurred approximately 10% of the time, while the vaccinated group experienced Covid contraction 90% of the time. Contrary to expectations, the vaccine appears to be ineffective in this testing scenario and, in fact, demonstrates an inverse effectiveness, seemingly elevating the risk of Covid contraction in this group. The adverse effects observed in this group surpass even those experienced by the already negatively affected LGBTQ group.

### Inferential Results

```{r}
chisq.test(tableDrug)
chisqtestGC(tableDrug)
drugFisher <- fisher.test(tableDrug)
drugFisher
```

The results of the Chi-Squared test reveal a highly significant finding, with a P value of $2.842\times10^{-9}$—considerably smaller than the commonly used threshold of 0.05. This small P value suggests a significant difference in the data. Additionally, the Chi-Squared statistic, standing at 35, surpasses the degrees of freedom (one) for this dataset. In this group, the likelihood of an unvaccinated individual contracting Covid is approximately 1 in 8.

### Confidence Interval

```{r}
oddsDrug = drugFisher$estimate/(drugFisher$estimate +1)*100
print(oddsDrug)
drugLow=(drugFisher$conf.int[1]/(drugFisher$conf.int[1]+1))*100
drugHigh=(drugFisher$conf.int[2]/(drugFisher$conf.int[2]+1))*100
cat("\n","Condifence interval for Drug Users: ",drugLow," to ",drugHigh)
```

The data shows that the confidence interval is 4-21%, which is far lower than 50%. This means we can say that  this is also not a very good vaccine.

# Overall Results and Conclusions

Upon examining all categories, the rejection of the Null Hypothesis is evident, albeit for diverse reasons. Certain groups demonstrated favorable outcomes where the vaccine proved effective, lowering the risk of contracting Covid. Conversely, other groups seemed to face an elevated risk upon receiving the vaccine. The exact cause for these disparities remains unclear from the data. An important point in the data, heterosexual men and women seem to experience benefits from the vaccine, placing them in a lower-risk category. In contrast, the LGBTQ and Drug User populations exhibit notably adverse outcomes from the vaccine.

## Males

Analysis of the data shows that the male group exhibited the most positive outcomes from the vaccine. Both percentage charts and bar charts indicate that men are less likely to contract Covid if they receive the vaccine. Statistically significant differences, supported by small P-Values and high Chi-Squared statistics, lead us to reject the Null Hypothesis in favor of the alternative, affirming the effectiveness of the vaccine in this group.

## Females

Examination of the data demonstrates positive outcomes for the female group from the vaccine. Similar to males, percentage charts and bar charts show a reduced likelihood of contracting Covid for women who receive the vaccine. Despite being less effective than in the male group, statistical significance, as reflected in small P-Values and high Chi-Squared statistics, allows us to reject the Null Hypothesis, confirming the vaccine's effectiveness in this group.

## LGBTQ

Data analysis reveals negative outcomes for the LGBTQ group from the vaccine. Both percentage charts and bar charts indicate an increased likelihood of contracting Covid for LGBTQ individuals who receive the vaccine. The statistically significant differences, with small P-Values and high Chi-Squared statistics, lead to rejecting the Null Hypothesis, emphasizing that the vaccine is not effective in this group and may even elevate their risk.

## Drug Users

Examination of the data for the Drug User group exposes the least favorable outcomes from the vaccine. Both percentage charts and bar charts highlight a higher likelihood of contracting Covid for drug users who receive the vaccine. The statistically significant differences, marked by small P-Values and high Chi-Squared statistics, lead to rejecting the Null Hypothesis. The vaccine proves ineffective in this group and appears to place them at an elevated risk, surpassing even the heightened risk observed in the LGBTQ category.